\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-24 tex/paper.tex
update notations

We consider a family of parameterized densities $p_\theta(\mathbf{x})$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p_{z(\theta)}(\mathbf{z}) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p_{z(\theta)}$ may itself depend on $\theta$ in some intricated way.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta)
\end{equation}
such that the likelihood $p_\theta(\mathbf{x})$ can be rewritten as
\begin{equation}\label{eqn:p_x_sim}
    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p_{z(\theta)}(\mathbf{z}) \mu(d\mathbf{z}),
\end{equation}
where $\mu$ is a probability measure.
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r$, our goal is the inference of the parameters
of interest $\theta^*$ that minimize the divergence between $p_r$ and
the modeled data distribution $p_\theta$ induced by $g(\cdot;
\theta)$ over $\mathbf{z}$. That is,
\begin{equation}
    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r, p_\theta),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\mathbf{z}$. More specifically,
a generative model $g(\cdot; \theta)$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r \parallel p_\theta)$ between $p_r$ and $p_\theta$.

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r, p_\theta)$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r} [d(\mathbf{x};\phi)]  \nonumber \\
                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p_{\hat{\mathbf{x}}}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p_\theta} [d(\tilde{\mathbf{x}};\phi)]
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} + (1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim p_r$ and $\tilde{\mathbf{x}} \sim p_\theta$.

@@

Evolution strategies~\citep{2011arXiv1106.4487W} and variational optimization~\cite{2012arXiv1212.4507S} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta)] = U(\psi),
\end{equation}
where $q_\psi$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta)] \nonumber \\
    &= \nabla_\psi \int f(\theta)  q_\psi(\theta)  d\theta \nonumber \\
    &= \int f(\theta) \nabla_\psi q_\psi(\theta)  d\theta \nonumber \\
    &= \int \left[ f(\theta) \nabla_\psi \log q_\psi(\theta) \right]  q_\psi(\theta)  d\theta \nonumber \\
    &= \mathbb{E}_{\theta \sim q_\psi(\theta)} [f(\theta) \nabla_\psi \log q_\psi(\theta)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log
q_\psi(\theta)$ of the proposal is known and that one can evaluate
$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

@@

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r$, simulator $g$.\\
        {\it Outputs:} proposal distribution $q_\psi(\theta)$, such that $p_r \approx p_\psi$.\\
        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q_\psi \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r, \theta_m \sim q_\psi, \mathbf{z}_m \sim p_{z(\theta_m)}, \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \theta_m \sim q_\psi,  \mathbf{z}_m \sim p_{z(\theta_m)} \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
parameters of $g$ and $p_\theta$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_{z(\theta)}}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
\end{equation}
for $\theta_m \sim q_\psi$ and $\mathbf{z}_m \sim p_{z(\theta_m)}$ and .
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \theta \sim q_\psi, \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
More formally, the learned model $p_\psi(\mathbf{x})$ therefore corresponds to the marginal distribution
$\int q_\psi(\theta) p_\theta(\mathbf{x}) d\theta$ of the generated data.
\glnote{Connection with variational bayes?}

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q_\psi$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] + \gamma H(q_\psi)
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.
\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}

>>>>

We consider a family of parameterized densities $p(\mathbf{x}|\theta)$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p(\mathbf{x}|\theta) = \int p(\mathbf{x}|\mathbf{z},\theta) p(\mathbf{z}|\theta) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p(\mathbf{z}|\theta)$ may itself depend on $\theta$ in some intricated way.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p(\mathbf{x}|\mathbf{z},\theta)$, as
specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p(\mathbf{x}|\theta) \equiv \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta)
\end{equation}
such that the likelihood $p(\mathbf{x}|\theta)$ can be rewritten as
\begin{equation}\label{eqn:p_x_sim}
    p(\mathbf{x}|\theta) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}|\theta) \mu(d\mathbf{z}),
\end{equation}
where $\mu$ is a probability measure.
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
e.g. as when $g$ is specified as a computer program.

Given some observed data $\{ \mathbf{x}_i | i=1, \dots, N \}$ drawn from the
(unknown) true distribution $p_r(\mathbf{x})$, our goal is the inference of the parameters
of interest $\theta^*$ that minimize the divergence between $p_r(\mathbf{x})$ and
the modeled data distribution $p(\mathbf{x}|\theta)$ induced by $g(\cdot;
\theta)$ over $\mathbf{z}$. That is,
\begin{equation}
    \theta^* = \arg \min_{\theta \in \Theta} \rho(p_r(\mathbf{x}), p(\mathbf{x}|\theta)),
\end{equation}
where $\rho$ is some distance or divergence.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\mathbf{z}$. More specifically,
a generative model $g(\cdot; \theta)$ is pit against an adversarial
classifier $d(\cdot; \phi):\mathbb{R}^d \to [0,1]$ with parameters $\phi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\mathbf{z}; \theta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\theta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\theta)$.

As thoroughly explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W(p_r(\mathbf{x}), p(\mathbf{x}|\theta))$ by
replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \phi) : \mathbb{R}^d \to \mathbb{R}$. Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\phi$ and $\theta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,& \mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]  \nonumber \\
                  & + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\phi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& -\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)]
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)$.

@@

Evolution strategies~\citep{2011arXiv1106.4487W} and variational optimization~\cite{2012arXiv1212.4507S} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
\end{equation}
where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\psi)$ is differentiable with respect to $\psi$, and using the log-likelihood
trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] \nonumber \\
    &= \nabla_\psi \int f(\theta)  q(\theta|\psi)  d\theta \nonumber \\
    &= \int f(\theta) \nabla_\psi q(\theta|\psi)  d\theta \nonumber \\
    &= \int \left[ f(\theta) \nabla_\psi \log q(\theta|\psi) \right]  q(\theta|\psi) d\theta \nonumber \\
    &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta) \nabla_\psi \log q(\theta|\psi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log
q(\theta|\psi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\theta})$ for any $\theta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, for scaling invariance and
robustness to noisy gradients).

@@

    \begin{flushleft}
        {\it Inputs:} observed data $\{ \mathbf{x}_i \}_{i=1}^N \sim p_r(\mathbf{x})$, simulator $g$.\\
        {\it Outputs:} proposal distribution $q(\theta|\psi)$, such that $p_r(\mathbf{x}) \approx p(\mathbf{x}|\psi)$.\\
        {\it Hyper-parameters:} The number $n_{\text{critic}}$ of training iterations of $d$; the size $M$ of a mini-batch; the gradient penalty coefficient $\lambda$; the entropy penalty coefficient $\gamma$.
    \end{flushleft}

    \label{alg:avo}
    \begin{algorithmic}[1]
        \State{$q(\theta|\psi) \leftarrow \text{prior on $\theta$ (with differentiable and known density)}$}
        \While{$\psi$ has not converged}
            \For{$i=1$ to $n_{\text{critic}}$} \Comment{Update $d$}
                \State{Sample a mini-batch $\{\mathbf{x}_m \sim p_r(\mathbf{x}), \theta_m \sim q(\theta|\psi), \mathbf{z}_m \sim p(\mathbf{z}|\theta_m), \epsilon_m \sim U[0,1] \}_{m=1}^M$.}
                \For{$m=1$ to $M$}
                    \State{$\tilde{\mathbf{x}}_m \leftarrow g(\mathbf{z}_m; \theta_m)$}
                    \State{$\hat{\mathbf{x}}_m \leftarrow \epsilon_m \mathbf{x}_m + (1 - \epsilon_m) \tilde{\mathbf{x}}_m$}
                    \State{$U_d^{(m)} \leftarrow d(\tilde{\mathbf{x}}_m; \phi) - d(\mathbf{x}_m; \phi) + \lambda (|| \nabla_{\hat{\mathbf{x}}_m} d(\hat{\mathbf{x}}_m; \phi) ||_2 - 1)^2 $}
                \EndFor
                \State{$\phi \leftarrow \text{Adam}(\nabla_\phi \tfrac{1}{M} \sum_{m=1}^M U_d^{(m)})$}
            \EndFor
            \State{Sample a mini-batch $\{ \theta_m \sim q(\theta|\psi),  \mathbf{z}_m \sim p(\mathbf{z}|\theta_m) \}_{m=1}^M$.}  \Comment{Update $q_\psi$}
            \State{$\nabla_\psi U_g \leftarrow \tfrac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m)) \nabla_\psi \log q_\psi(\theta_m)$}
            % \State{$\mathbf{F}(\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi \log q_\psi(\theta_m) \nabla_\psi \log q_\psi(\theta_m)^{\top}$}
            \State{$\nabla_\psi H(q_\psi) \leftarrow \tfrac{1}{M} \sum_{m=1}^M  \nabla_\psi q_\psi(\theta_m) \log q_\psi(\theta_m)$}
            \State{$\psi \leftarrow \text{Adam}(\nabla_\psi U_g + \gamma \nabla_\psi H(q_\psi))$}
        \EndWhile
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$ and .
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p(\mathbf{x}|\psi) \equiv \theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta), \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q(\theta|\psi)$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r(\mathbf{x})$.
More formally, the learned model $p(\mathbf{x}|\psi)$ therefore corresponds to the marginal distribution
$\int q(\theta|\psi) p(\mathbf{x}|\theta) d\theta$ of the generated data.
\glnote{Connection with variational bayes?}

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyperparameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.
\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}
